---
title: Decision making under uncertainty
author: Vlad Aluas
date: '2021-02-21'
slug: decision-making-under-uncertainty
categories: ["Opinion"]
tags: ["Monte Carlo", "decision making", "uncertainty", "chance"]
thumbnail: /images/chance-logo.png
draft: true
---

**Insert Lead paragraph here.**

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", collapse = FALSE, out.width = "100%")

library(tidyverse)
library(scales)
```

------------------------------------------------------------------------

People engage in decision making on a daily basis and although most of our decisions seem to derive from logical arguments, the truth is that exactly the opposite is true. Most of the decisions we make are due to [biases](https://en.wikipedia.org/wiki/List_of_cognitive_biases) and [heuristics](https://en.wikipedia.org/wiki/Heuristic) our minds have developed to respond to the environment around them faster.

These tendencies have been first described and studied by [Amos Tversky](https://en.wikipedia.org/wiki/Amos_Tversky) and [Daniel Kahneman](https://scholar.princeton.edu/kahneman/home) and have been explained in Kahneman's excellent book *Thinking fast and slow*. I highly recommend the book since it is an excellent guide in where our minds tend to be error prone and mislead us. I will explain in brief however that our minds operate as two separate system called by Kahneman and Tversky *system 1* and *system 2*. 

*System 1* is the default system of the mind, the one used for everyday activities. These activities include walking, breathing, detecting danger and avoiding it, eating, doing routine activities, and all activities that do not involve heavy cognitive effort. *System 1* is emotional, automatic, fast and uses no extra energy from our energy hungry brains.

*System 2* is the exact opposite of *System 1*. It is rational, quite slow, and requires a large amount of energy and effort from our part. *System 2* allows you to analyse in detail every situation, however this is not always ideal. If a car would come speeding towards you, *System 2* would analyse the colour of the car, the brand and maybe the speed and approach vector, which is a nice thought exercise but might get hit by the car before *System 2* would act. *System 1* will not bother with all those details, it will just register car coming towards me with great speed, I need to move out of the way.

The problem when making decisions is that there are situations moderated or mediated by the aforementioned biases and heuristics in which *System 1* might make the decision and our minds might think it is actually *System 2*. Otherwise said, we might make decisions for subjective reasons and have a high degree of confidence in then believing they are objective.

In this article I would like to show some methods that can reduce these types of biases and help in engaging *System 2* in the decision making process.

# Decision Consequences

I would like to begin with a brief by pointing out that not all decisions we make have the same consequences, or affect us in the same way. Your decision to have a pretzel for breakfast instead of a salad might have consequences later in life if you repeat this behaviour multiple times. Otherwise, you are quite safe to do so from time to time.

Not the same can be said about a decision to bet all your savings on a race or a sports game. Considering your saving amount to a decent sum, your bet can leave you without anything you have saved or can increase your saving quite drastically.

So, as we can see consequences can vary quite a lot in the outcome type (positive / negative) and in their magnitude. The focus of this article will be to check how we can better understand the consequences of our decisions, try to avoid large negative outcomes and should it be the case seek out large positive ones.

# Distribution Types

In order to be able to continue, I want to explain that not all things are as predictable as they seem and this is because not all things are distributed equally.

What do I mean by that? Let's give some examples to make it clear.

Let's say you go to a festival (I know, I wish it as well) of around 10.000 people. Considering the people who tend to frequent such gatherings, I will guess that the average age of the people there is between 22 and 27 years old. For the purpose of our example I will say 24.5.

Now, think of the oldest person you have heard of, it can be anyone does not matter. Add their age to the calculation of the average. Does it make a bit difference? No, I does not. Even if you think of the oldest fictional person, Methuselah, with his age of 969 does not impact the average by much (I know some sumerian kings are supposed to have lived for thousands or tens of thousands of years, however they used a different counting system so we cannot be sure how they actually counted). So, let's see the impact of Methuselah on the average age.

$$m_{age} = \frac{10,000 \times 24.5 + 969}{10,001} = `r (10000 * 24.5 + 969) / 10001`$$

Not such a big impact is it.

For the second example of a distribution, I would like to see the average net worth of the crowd. I will assume that since the people at this festival are quite young, most of them do not have a lot o noteworthy possessions as of yet, however to compensate for those who do, I will say the average is around \$65K. Now, following the logic we used above, if you think of the richest man you know of and add him or her (Oprah can just as easily skew the data) to the crowd, how much will the average change?

I'm thinking of Elon Musk although at the time of writing this article he is work a measly \$169B as opposed to Jeff Bezos' \$193B, however I'm a big fan of Musk so let's check with his net worth.

$$m_{net\_worth} = \frac{10,000 \times 65,000 + 169,000,000,000}{10,001} = \$16,963,304$$ Quite a difference, right? I'd say so.

### Normal Distribution

In the first example the data is not that impacted that strongly because it follows a [normal distribution](https://en.wikipedia.org/wiki/Normal_distribution). A normal distribution is a phenomenon observed by [Carl Gauss](https://en.wikipedia.org/wiki/Carl_Friedrich_Gauss) in which the characteristics of a population (of a variable to be more precise) tend to cluster towards the average value and the further you go from the average value, the less likely it is to happen.

Basically, if we have a set of data that follows a normal distribution, we can be quite certain that outliers, or values that will significantly impact our outcome are highly unlikely to happen. In out example, you cannot point to a single person that can impact the average height of the festival, only if you find someone who is several kilometres tall.

These types of data usually are the ones that measure natural phenomena, as mentioned earlier. They can include but not be limited to:

-   height
-   weight
-   age
-   reaction times
-   average time spent playing video games

### Power Law Distribution

The other effect we can observe is called a power law and sometimes a [Pareto Distribution](https://en.wikipedia.org/wiki/Pareto_distribution), especially in finance, or the 80-20 rule. The last one is the best known and is usually formulated as 20% of individuals control 80% of the assets.

As we could see in the example above, in such a distribution, just one observation can impact everything significantly and that is why we need to take the type of distribution into consideration when taking a decision. Not taking this into consideration we have a lot of people calling themselves actors, writers or singers and not what they actually do for a living, most likely bartenders, waiters or other job unrelated with their self-reported career.

This is not meant as an insult for the people who do this, it is just to point out that in some fields, especially the ones that are scalable (once you create something it is easy to replicate without any effort from your part) this distribution tends to be very skewed and a few people control most of the resources.

### Distributions in Decision Making

The most important thing to remember in regards to distributions is that if we have to make a decision in regards to a normal distribution, we can be pretty sure that the average of the distribution is quite indicative of what is happening. We can use a sample without any problem. Does not matter whom do we pick from a crowd to measure the average height, even if we get the average height wrong, it will be off by just a small margin.

This cannot be said about a power law distribution. One individual is nothing like each other, you either have dwarves or giants and if you take a sample of dwarves you cannot measure the impact of a giant, and therefore an average of any measure from a sample of such a distribution is useless.

So, how do we distinguish between the two kinds of distributions in the real world and how can we apply this knowledge in decision making? For this there is no straightforward answer, however, there are some methods that can help in evaluating the impact of different decisions.

In the next section I will describe a problem, I will explain my reasoning in adding and analysing variables to it and I will show how you can use a [Monte Carlo method](https://en.wikipedia.org/wiki/Monte_Carlo_method#:~:text=Monte%20Carlo%20methods%2C%20or%20Monte,might%20be%20deterministic%20in%20principle.) to make sure that you can take the best decision out of all of them. Or the one that is best suited for your needs, I should say.

# The problem

Let us imagine that you would like to choose a career. You are not really sure what you would like to choose because you don't have a particular career for which you have a strong calling, however there are some that seem decent to you:

-   The first option would be doctor. You have seen a lot of medical procedurals and it looks great and exciting, however it would be even better if you can set up your own practice and be the master of your own destiny. Also it would be better paid
-   The second option would be to work in IT. It still seems like a great job because you can have a lot of puzzles with which to play and if you mess something up you don't endanger someone's life.
-   The third option is to disregard all safe plans and go for something that will let you retire early. After all, you have seem many people win big at poker of blackjack and they are far from you cognitive prowess, therefore it might be riskier, however, if anyone has a chance of doing it, it's you.

However, all these options seem good and you can't really choose out of any of them. So, how do you decide with which career you would end up better at retirement? For the sake of this example and to make it easy to understand and reproduce, we will only take into account the amount of money you will earn over your career, however using the same principles you can calculate as many variables as you like.

# The method

Since you are an amazing fellow, you can assume that you will not only outperform your peers in gambling, but you will do so even if you are a dentist or a programmer, so we will eliminate this from the equation since it is a constant. So in order to answer the question we will check how an average person performs the task.

Since one example cannot tell us much about what might happen, we will set up some conditions for each of the career paths and generate some random numbers that abide by each condition.

This will simulate our outcome at retirement if we were to live multiple lives and work in each domain for some of those lives. After this we can simply check the results and see which one of the careers would give us the better chance for a retirement.

## Coding examples

In order to better asses the situation, I will use some code. The code is written in *R*, however it can be ignored completely if you do not want to bother with it. I will try my best to comment the code properly so you can follow even if you are not familiar with *R*. If you want to learn more about this particular programming language, I would recommend [R for Data Science](https://r4ds.had.co.nz/), a brilliant book that is the perfect starting place.

Let's start coding. At first I would like to set up the variables that I will use in this article. I will explain later why I use these values, however I will use them repetedly so I will like to set them up now avoid changing them later in multiple places.

```{r variables, include = FALSE}
# Medical Salaries
dr_sal_mean <- 21100

# IT Salaries
it_sal_med_high <- 225000 # Median Salary for Managers and Directors in IT
it_sal_med_low  <- 92809  # Median Salary for the everyone else working in IT 
```

Next I will like to write some custom function that will allow us to focus on the results, not on endless lines of code.

```{r functions, include = FALSE}
# Graphical function to avoid writing the same lines over and over
histogram <- function(.df){
  
  .df %>% 
    tibble() %>% 
    ggplot(aes(.)) +
    geom_histogram(colour = "white", fill = "#2484cc") +
    labs(x = "Income", y = "") +
    scale_x_continuous(labels = label_dollar(prefix = "€"))
}

# Just a simple function to format the Romanian currency so I can used stored variables.
ron_format <- function(.x){
  require(scales)
  
  dollar_format(prefix = "", suffix = " RON")(.x)
}
```

# The careers

In the next section I will describe each career, the variables that I will take into account and I will expose the code I will use to do so, if anyone wants to replicate it. The code is written in *R*.

## The Doctor (of the medicine not Who variety)

I all societies doctors are well respected and well paid. However, not all doctors are paid the same. We can assume that a neurosurgeon is paid better than a internist although not as well as a plastic surgeon.

I have found a data source that seems to have done a pretty good job in assessing the average salary for a doctor. The information was taken from the [salaryexplorer.com](http://www.salaryexplorer.com/salary-survey.php?loc=178&loctype=1&job=13&jobtype=2) website. The salary is expressed in the local currency of Romania (RON) and the salaries are expressed before tax. These two impediments are nothing too difficult as we can easily translate everything into Euro, and extract the taxes.

According to the National Bank of Romania, the conversion rate is 1 RON for 0.20 €.

In regards to the taxes, the Romanian government imposes taxes on the on the employee as follows:

-   Social Security: 25%
-   Health Insurance: 10%

The remaining 65% is considered income and that is taxed with 10%.

Let's start by extracting the taxes out of the average salary for doctors of `r ron_format(dr_sal_mean)`.

$$ m_{salary(RON)} = (`r dollar_format(prefix = "")(dr_sal_mean)` - \frac{25}{100} \times `r dollar_format(prefix = "")(dr_sal_mean)` - \frac{10}{100} \times `r dollar_format(prefix = "")(dr_sal_mean)`) \times \frac{90}{100} = `r dollar_format(prefix = "")(dr_sal_mean - 0.25 * dr_sal_mean - 0.1 * dr_sal_mean)` \times \frac{90}{100} = `r dollar_format(prefix = "")((dr_sal_mean - 0.25 * dr_sal_mean - 0.1 * dr_sal_mean) * 0.9)`$$

Now we can easily convert everything into Euro.

$$ m_{salary(€)}  =  `r dollar_format(prefix = "")((dr_sal_mean - 0.25 * dr_sal_mean - 0.1 * dr_sal_mean) * 0.9)` \times 0.2 = `r dollar_format(prefix = "")(((dr_sal_mean - 0.25 * dr_sal_mean - 0.1 * dr_sal_mean) * 0.9) * 0.2)`$$

With this in mind, our average income on a monthly basis, if we were to live 10000 times would be distributed something like this.

```{r dr_monthly_salary, message = FALSE}
dr_sal_euro <- ((dr_sal_mean - dr_sal_mean * 0.25 - dr_sal_mean * 0.1) * 0.9) * 0.2

dr_sal_dist <- rnorm(10000, mean = dr_sal_euro, sd = 600)

dr_sal_dist %>% 
   histogram() +
   labs(title = "Monthly Income Distribution",
        subtitle = "Medical Care Workers")
```

Now, we will just need to calculate for how long does a doctor work and some other costs of being a doctor. In Romania, medical school is usually 7 years, and you usually need to spend extra 3 to 5 years for residency, although that is usually paid so I think we can just include the 7 years of study. If you finish high-school at 19 and you spend 7 years in medical school, you can start working and 26.

If we want to retire at 65 and, that would give us 39 years of work or 468 months of works. So, all that we have to do is create a distribution with the mean salary after tax and multiply it by the number of months.

```{r dr_lifelong_salary, message = FALSE}
dr_sal_dist_total <- dr_sal_dist * 468

dr_sal_dist_total %>% 
   histogram() +
   labs(title = "Lifelong Income Distribution",
        subtitle = "Medical Care Workers")
  
```

Not bad, right?. Now let's see how much would someone in IT make.

## Part of the IT Crowd

Romania has one of the most active IT sectors in Eastern Europe and it is expanding continuously which is a good thing because one will not have problems ending up in an IT job if he wants to. We will just need some data on the salaries of the IT crowd. A beautiful analysis on 1,267 salaries can be seen [here](https://www.averagesalarysurvey.com/romania).

The sample is large enough to give us a good estimate of what the population actually looks like. We can see that the data is a bit skewed and slightly more people are just under the mean when it comes to income. This is due to the fact that managers and directors have very high salaries and they move the average up. This can be easily fixed, we just need to estimate what percentage of people working in IT are managers or directors and and account for that.

The first table gives us most of the information we need regarding the salary distribution in Romania. There we can see that most people, over 58% earn more than `r ron_format(it_sal_med_low)`. This is not exactly the median of the distribution however it is close enough so we will use this to approximate the shape of the distribution. The median is the value exactly in the middle of the distribution.

Once we have this, we will use it as an average value for a normal distribution that will contain most of the values in our population. This will be 85% percent of our lives, we will not be managers, therefore we will have salaries accordingly. In the rest of 15% of our 10,000 lives, we will be managers or directors so our salary will be accordingly.

```{r it_salary_distribution}
it_sal_low <- rnorm(8500, it_sal_med_low, 25000)
it_sal_high <- rnorm(1500, it_sal_med_high, 50000)

it_sal_dist <- c(it_sal_low, it_sal_high)

# We want to make sure the mean of our distribution if not far off from the actual distribution
mean(it_sal_dist)
```

We can see the average salary of our distribution is quite close to the one from the website, so we can say that our distribution closely resembles the sampled population from the site. Salary distribution looks like below:

```{r it_yearly_salary, message = FALSE}
it_sal_dist %>% 
  histogram() +
  labs(title = "Yearly Income Distribution",
       subtitle = "IT Workers") + 
  scale_x_continuous(labels = label_dollar(prefix = "", suffix = "RON"))
```

Not very normal, is it? There are a few people that have huge salaries compared to the rest of the population. So, if you manage to land yourself a management role in IT you're all set. If not, let's see how this translates in EURO and in lifetime earnings.

The first thing would be to transform everything into net profit.

At the time of this writing the IT workers were exempt from the income tax so the only two taxes required to be paid are:

-   Social Security: 25%
-   Health Insurance: 10%

Afterwards, we will convert it to Euro.

```{r it_sal_dist_net}
it_sal_dist_net <- it_sal_dist - (it_sal_dist * 0.25) - (it_sal_dist * 0.1)

it_sal_dist_net_eur <- it_sal_dist_net * 0.2
```

Now we have the yearly salary. So, for how long can someone work in this industry?

Most people start while studying at university, however most of those positions are either very low paying or are unpaid internships so I will eliminate them. On the other hand, due to this, most people can find work right after finishing their studies. Assuming as we did previously that someone finishes high school at 19, they can study for 3-4 years, depends if you are studying in the Bologna system or not, you can start work at 22. However, since there are a lot of people working in IT who do not have specialisations in IT, I will consider that the most typical person will start at 25. This will account for those people as most people not from IT might start later.

Considering that my goal is to retire at 65, I have 40 years of work ahead of me, so all I need to do is multiply everything by 40.

```{r it_sal_lifelong_income, message = FALSE}
it_sal_dist_net_eur <- it_sal_dist_net_eur * 40

it_sal_dist_net_eur %>% 
  histogram() +
  labs(title = "Lifelong Income Distribution",
       subtitle = "IT Workers")
```

This looks decent enough. Not as good as a doctors, however not bad at all. Now, I would like to take into account one more part.

What if I develop an app and make it big with it?

Honestly, I checked for relevant statistics regarding the success of an app. You can read a very interesting summary in this [article](https://venturebeat.com/2017/09/24/your-chances-of-making-a-successful-mobile-app-are-almost-nil/). From here and other sources it shows that to actually develop an app you need an initial investment of at least €25,000 and make it a full time job. Afterwards you need to invest in marketing and spend a lot of time adjusting ASO.

This is kinda outside the scope of what I want to do so I will not include it in my final outcome.

## Poker player

Now let's compute the outcome of ten thousand lives as a online poker player. For this we will need some additional information:

-   What is the probability to make it as an professional online poker player
-   How much or how little does an online poker player make in a year
-   How is that money distributed

These assumption can be estimated in multiple ways, although most of them are time consuming and the technical methods used are outside the scope of this article. However, I will briefly mention them if you are interested in such an endeavour.

[Statname.net](https://statname.net/) is one of the multiple sites that offer statistics regarding player earnings on different platforms and strategies. You can get the data from there on the earnings of a player, the number of hands and compute how much he would have earned if you account for the blinds. Then you can set a cut-off point in which an earning is considered a living wage and you can compute the percentage of professional players. This will be your probability. The same can go for the other measures.

Since getting the data can be a bit time consuming and this being just an example, I have decided to base my computations on the following [blog](https://www.blackrain79.com/). You can get a lot of useful information about online poker from here and I found it very useful, my understanding of poker being the as basic as it could be. From what I read here, and discussed this with people who know about the subject way more than I do, apparently one of the most successful strategies in online poker is to win blinds not large hands. The tables are also divided by the size of the blinds as follows:

-   NL2 (also known as 2NL) = ¢1 / ¢2 blinds
-   NL5 (also known as 5NL) = ¢2 / ¢5 blinds
-   NL10 (also known as 10NL) = ¢5 / ¢10 blinds
-   NL25 (also known as 25NL) = ¢10 / ¢25 blinds
-   NL50 (also known as 50NL) = ¢25 / ¢50 blinds
-   NL100 (also known as 100NL) = ¢50 / \$1 blinds

Now, how do we transform this into earnings? The formula is simple as seen below:

$$ (winning\_big\_blinds\_chance \times big\_blind\_value) \times \frac{played\_hands}{100}  $$

Let's compute some and example. From what I could gather a professional player will play around 100.000 hands a month. So from here on I will use this as the number of hands a professional player will play so we can estimate the winnings. With this in mind, let's see how much will a professional player make from a NL2 table with a 30% win rate. That will translate into the following formula:

$$ (30 \times 2) \times \frac{100000}{100} = ¢60000$$

We need to keep in mind that those are cents, adjusting to dollars, we see that you can earn around \$600 a month playing low stakes poker, which is not bad. We also need to take into account that people can play multiple tables online and apparently most people do. So the probabilities that I am going to use here can be found in this [blog post](https://www.blackrain79.com/2014/06/good-win-rates-for-micro-and-small_6.html) and are going to be just the ones for 1-8 tables since I already established that I will play around 100.000 hands a month so increasing the number of tables to 18+ cuts time but lower my attention and therefore the winning rate. If I want to maximise the number of hands, or increase the hourly payment I would increase the number of tables.

I will compute the average for each type of table, however I will transform the cents into dollars directly

$$ NL2   = (10 \times \frac{2}{100}) \times \frac{100000}{100} = \$`r (10 * 2/100) * 1000`$$

$$ NL5   = (6 \times \frac{5}{100}) \times \frac{100000}{100} = \$`r (6 * 5/100) * 1000`$$

$$ NL10  = (4 \times \frac{10}{100}) \times \frac{100000}{100} = \$`r (4 * 10/100) * 1000`$$

$$ NL25  = (3 \times \frac{25}{100}) \times \frac{100000}{100} = \$`r (3 * 25/100) * 1000`$$

$$ NL50  = (2 \times \frac{50}{100}) \times \frac{100000}{100} = \$`r (2 * 50/100) * 1000` $$

$$ NL100 = (2 \times 1) \times \frac{100000}{100} = \$`r (2 * 1) * 1000` $$

Now that we have some averages, I would like to see how much can the best players make using the same estimates from the articles.

$$ NL2   = (30 \times \frac{2}{100}) \times \frac{100000}{100} = \$`r (30 * 2/100) * 1000`$$

$$ NL5   = (10 \times \frac{5}{100}) \times \frac{100000}{100} = \$`r (18 * 5/100) * 1000`$$

$$ NL10  = (12 \times \frac{10}{100}) \times \frac{100000}{100} = \$`r (12 * 10/100) * 1000`$$

$$ NL25  = (10 \times \frac{25}{100}) \times \frac{100000}{100} = \$`r (10 * 25/100) * 1000`$$

$$ NL50  = (9 \times \frac{50}{100}) \times \frac{100000}{100} = \$`r (9  * 50/100) * 1000` $$

$$ NL100 = (8 \times 1) \times \frac{100000}{100} = \$`r (8 * 1) * 1000` $$

We can see that this is not anywhere near a normal distribution. By the definition of professional players, we know that they are the ones that actually earn enough so that they can survive of this income. This leaves us with a skewed distribution. I think we can safely assume that most professional players are average players and the elite that earn as much as we have shown here are a small percent 10%. With that said, we can start creating the population of online poker players.

Another assumption is that professional poker players will not spend all their time in NL2, NL5 and NL10 games, will most likely spend it on something that will provide more income. This is also consistent with the living wage assumption. I know \$750 might not seem much in a month for most Western democracies, however, it is quite a decent income in second and third world countries. So, how are we going to do this?

Simple, we will create two distributions that will reflect the average player and the elite player. The first one will be 90% of the population and the second one will be the other 10%

```{r}
avg_player <- rnorm(9000, 1400, 300)
elt_player <- rnorm(1000, 4000, 1000)
poker_players <- c(avg_player, elt_player)
```

```{r}
poker_players %>% 
  histogram() +
  labs(title = "Monthly Income Distribution",
       subtitle = "Professional Poker Players")
  
```

This distribution seems legit for a power law distribution. And now, for the final touch, we will need to add the 10% chance to become a professional player. We will do this by using the list we have created earlier. For this we will create a binomial distribution with a 10% probability and a size 1. Basically we can create a random string of values, either 1 or 0 and about 10% are one. Let's see it in practice.

```{r}
success_prob <- rbinom(n = 10000, size = 1, prob = 0.1)
print(sum(success_prob) / 10000)
```

Perfect, `r scales::percent_format(accuracy = 0.01)(sum(success_prob) / 10000)` is exactly what we would expect.

Now all we need to do is multiply the two lists. When a value in the second list is 0 it will transform the earnings to 0, so we did not make it as a professional poker player. When it is 1 the result will not be affected.

```{r}
poker_players_final <- poker_players * success_prob
```

Now that we painfully and to the best of our knowledge created the populations, let's proceed in analysing the data.
